Fractals are Cool

### 2020-12-08 paper/paper_papaja.rmd
update again

abstract: |   
 Fractals can be described as geometric objects which contain self similarity and scale invariance. Fractal objects and fractal properties are pervasive in nature.  Many processes which involve replication or growth can be modeled with fractal mathematics. The formalized mathematics of fractals has been used as an analog for complexity, an important property related to human cognition. Studies have shown that human perception has evolved to process fractals fluently, as well as exhibit aesthetic preferences for fractals. The experience of fractals in a number of domains have proven to be therapeutically useful. Previous studies have used independent and static fractal images to investigate the relationship between, fractals, complexity and human perception. This study employs a Dwell Time Paradigm to present self paced slide shows of dynamic fractals in an unfolding process of iterative growth and decay. Results of measured looking or dwell times show that participants prefer viewing fractals as they grow through phases of increasing complexity compared to decay phases of decreasing complexity. Longer dwell times for higher complexity may indicate that human perception is captured by the surprising or novel nature of a fractal’s iterative growth process.

 This study is the first to use a novel presentation of dynamic fractal processes to explore fractal fluency and aesthetic preference. Results from 16 participants prove that dynamic fractals can be integrated into the well established dwell time paradigm. These findings can be used to develop new techniques for utilizing fractals as therapeutic tools.
  <!-- https://tinyurl.com/ybremelq -->

keywords          : "keywords"
wordcount         : "X"

@@

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

---

```{r setup, include = FALSE}

@@

library(janitor)
library("papaja")
r_refs("r-references.bib")
# Sys.setenv(LANG = "en")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

# Introduction
  Fractal is a term originally derived to describe the “roughness,” (@mandelbrot1989fractal) of objects, surfaces, or textures.  Fractal properties are pervasive throughout nature (@losa2009fractal) and can be identified in the texture of clouds and mountains, in the branching pattern of trees and lightning, in the meandering shape of rivers and blood vessels.  These fractal properties are identified by their self-similarity and scale invariance. Self-similarity implies that a part of the structure maintains the same structure or pattern as the whole.  For example, if one were to zoom in on a small stretch of coastline, the jagged edges would look similar to the jaggedness of the coastline from a broader view.  Likewise, a small section of a fern is nearly identical to the branching shape of the entire fern.

```{r, fig.show = "hold", out.width =  "30%",fig.align = 'left'}
knitr::include_graphics(here::here("paper","paper_images","coast.jpg"))
knitr::include_graphics(here::here("paper","paper_images","fractalfern.png"))
knitr::include_graphics(here::here("paper","paper_images","spooky.png"))

```

  Scale-invariance is a related mathematical property which can be applied to objects or laws that do not change if scales of length, energy, or other variables, are multiplied by a common factor, and thus represent a universality. (wiki, scale invariance)  Expressed more intuitively, scale invariance implies that it does not matter at which scale one views and object or observes a process, because the object or process contains a commonality at all scales.

@@

  In addition to inventing the term fractal, Mandelbrot developed a mathematics and geometry of fractals.  He devised a way to model and generate complex systems with a simple mathematical rule that could be run repeatedly or iterated on itself.  These iterative processes essentially use a rule to take an initial condition or state, change or modulate it, then put the resulting updated state back into the rule to generate another new state.  This can be repeated infinitely and often results in high complex objects or systems.  Fractal mathematics have continued to be developed and utilized in mathematics, physics, biology, chemistry, and engineering.  A recent application of fractal mathematics is utilized in computer generated images and models.  By starting with a simple algorithm, a mathematical equation or rule, graphic artists and programmers can create highly complex models of natural systems.

@@

  Two premiere examples of these mathematical or statistical fractals can be seen in the Koch Curve and the Cantor Set.  In each case an image is created by starting with a single line segment and a rule for changing that line segment.  The cantor set can be visualized by displaying each cycle of the process of iteration together.  First a single line segment is displayed.  The Cantor Set rule divides the segment in three equal sections and removes the middle section, leaving two line segments each one third of the original length and separated by a space equal to one third of the original line segment.  The two separated line segments of the 2nd order iteration can be displayed beneath the original first order line segment.  The rule, dividing and removing a portion of the line segment is applied to the two new line segments.  Each of these line segments is divided into thirds and the middle third is removed. This results in four new smaller line segments spaced in a proportion equal to the previous two line segments.  These four line segments are again displayed under the two line segments.  This process can be repeated infinitely many times, dividing and removing segments at each iteration.  When displayed together, we are able to visualize the repeating pattern which relates to the simple unchanging rule.  This is an intuitive way to represent the properties of self-similarity, scale invariance, and iterative growth rules of fractals.  A similar example is the Koch Curve. Again we start with a single line segment, but define a new rule to change that line segment through each iteration.  Like the Cantor set, the line segment is divided into thirds, but instead of just removing the middle third which results in two new line segments, we bridge the middle third with two line segments that compose two sides of an isosceles triangle which spans the removed middle third of the original line.  This new segmented line of the 2nd order iteration appears as a line with a triangular bump or protrusion in the middle. The entire segment is composed of four connected line segments.  Again, by using the same algorithmic rule, divide a segment into thirds and span the removed middle third with an isosceles angle, and is applied to all four of the line segments created by the first iterative change.  Now each of these four line segments will include an angular bump or peak that spans their middle third.  These results in a set of 16 angular connected line segments.  Just as we did with the cantor set, and can do with all fractal growth iterations, we can continue applying the rule to the updated condition of the object.  This time, unlike the cantor set which displays all the iterations together, a single, new object is produced at each iteration.  We can visualize the growth process by displaying a few of these objects together.

```{r, fig.show = "hold", out.width =  "50%",fig.align = 'left'}
knitr::include_graphics(here::here("paper","paper_images","bars.png"))
knitr::include_graphics(here::here("paper","paper_images","star.jpg"))

```
  There are numerous examples of these types of geometric transformations which result in beautiful and striking constructions.  Some well known fractal geometries are known as, The Dragon Curve, Sierpinski Triangle, and Hilbert Curve.  These examples show the diversity of implementations of fractal properties in geometry.  The variety of initial conditions and algorithmic rules result in an astounding array of complex objects which have real world applications and instantiations.

```{r, fig.show = "hold", out.width =  "30%",fig.align = 'left'}
knitr::include_graphics(here::here("paper","paper_images","squigly.jpg"))
knitr::include_graphics(here::here("paper","paper_images","triangle.jpg"))
knitr::include_graphics(here::here("paper","paper_images","maze.jpg"))

```

@@

```{r, out.height =  "30%",fig.align = 'left'}
knitr::include_graphics(here::here("paper","paper_images","morefractals.png"))
knitr::include_graphics(here::here("paper","paper_images","square.png"))

```

@@

  To date there have been numerous research studies which explore the aesthetic preferences for fractals of different dimensions. Studies have included methods which present participants with a two alternative forced choice task of two fractals of differing dimension. Participants select the fractal image they prefer aesthetically and after viewing many pairs with differing dimensions, statistics are used to calculate the most preferred dimensionality range across participants. The overarching results of previous research have shown that humans, both adults and children prefer fractals within a range of dimensions that coincide with the fractal dimension that is found in natural objects and scenes, usually between 1.3 and 1.5.Remembering that fractal dimension is an analog for complexity, we can conclude that most humans prefer a moderate level of complexity compared to low or high complexity near the extremes. A secondary analysis of fractal fluency has shown that humans often also prefer fractal images to non-fractal images and that we prefer fractals which more closely resemble natural images to fractals which are geometrically exact. (@spehar2003universal) Extensions on these research findings have led to the use of fractal images in therapeutic applications.  Studies have shown some benefits of experiencing fractals, including physiological responses in EEG measurement of brain activity while viewing fractals (@hagerhall2015human), the reduction of stress using fractal art and architecture (@taylor2006reduction) and even in the experience of hearing fractal melodies (@beauvois2007quantifying). There appears to be a clear link between, the fractal’s representation of complexity, the connection with natural objects and processes, and the human mind.

@@

```{r, fig.show = "hold", out.width =  "30%",fig.align = 'left'}
knitr::include_graphics(here::here("paper","paper_images","blotches.png"))
knitr::include_graphics(here::here("paper","paper_images","moreblotches.png"))
knitr::include_graphics(here::here("paper","paper_images","deconstruction.png"))

```

@@

  Participants viewed self paced slideshows of fractal images in sequences of iterative growth and decay phases as well as a random ordering.  The time spent viewing each image was measured as a Dwell Time, following the procedures of Baldwin et. al. 2019.  A set of images were generated for each of fourteen different fractal types iterated from 4 to 9 times.  Each participant viewed a sequential slideshow and a random slideshow.  Sequential slideshows presented each fractal type through phases of growth and decay from the least complex to most complex iterations and back down from most complex to least complex.  Random slideshows presented the entire set of fractal images in random order.  6 participants viewed Sequential Slideshow A, starting with Fractal Type 1 to Type 14, and Random Slideshow A starting with Fractal Image 1 to Image 174.  The remaining 10 participants viewed Sequential Slideshow B, starting with Fractal Type 14 to Type 1, and Random Slideshow B starting with Fractal Image 174 to Image 1.  This counterbalancing was included to address the findings of previous dwell time studies in which participants dwell times systematically decrease through the slideshow.  In this way, the same fractal types were not always seen at the beginning or end of the slideshow.  This balancing was then reversed in analysis so that each fractal type and the corresponding dwell times could be analyzed for the entire set of 16 participants together.  Fractal images were generated using Wolfram Mathematica and slideshows were programmed using PsychoPy 3.

```{r, fig.show = "hold", out.height =  "25%",fig.align = 'left'}
knitr::include_graphics(here::here("paper","paper_images","task_outline.jpg"))
knitr::include_graphics(here::here("paper","paper_images","fractal_direction.png"))
```

##Participants
The study recruited 16 healthy young adults ages 16-35 through the UO SONA recruitment system.  All participants had normal or corrected to normal vision with no history of neurological damage.

##Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

@@

 A two-way ANOVA was conducted to examin the effect of dwell times on fractal presentation directions. The results indicated a significant interaction between the effects of fractal direction and dwell time, *F*(2,31.01) = 42.68, *p*<0.05). Simple main effects showed that growth conditions showed significantly longer dwell times compared to the decay or random directional conditions. 
```{r, fig.show = "hold", out.height =  "25%",fig.align = 'left'}

knitr::include_graphics(here::here("presentation_images","group1.png"))

```

  The following graph plots the dwell times over the course of the study for participant group 1. Group 1 and 2 where counterbalanced in fractal presentation order but show similar trends of decreasing dwell times as image number increases or the task itself progresses. Decreases in dwell times over time are likely attributed to fatigue or waning attention as the participants lose focus over time.      
```{r, fig.show = "hold", out.height =  "25%",fig.align = 'left'}

knitr::include_graphics(here::here("presentation_images","ovr_dwell_time2.png"))

```

>>>>

abstract: |
 Fractals can be described as geometric objects which contain self similarity and scale invariance. Fractal objects and fractal properties are pervasive in nature. Many processes which involve replication or growth can be modeled with fractal mathematics. The formalized mathematics of fractals has been used as an analog for complexity, an important property related to human cognition. Studies have shown that human perception has evolved to process fractals fluently, as well as exhibit aesthetic preferences for fractals. The experience of fractals in a number of domains have proven to be therapeutically useful. Previous studies have used independent and static fractal images to investigate the relationship between, fractals, complexity and human perception. This study employs a Dwell Time Paradigm to present self paced slide shows of dynamic fractals in an unfolding process of iterative growth and decay. Results of measured looking or dwell times show that participants prefer viewing fractals as they grow through phases of increasing complexity compared to decay phases of decreasing complexity. Longer dwell times for higher complexity may indicate that human perception is captured by the surprising or novel nature of a fractal’s iterative growth process. This study is the first to use a novel presentation of dynamic fractal processes to explore fractal fluency and aesthetic preference. Results from 16 participants prove that dynamic fractals can be integrated into the well established dwell time paradigm. These findings can be used to develop new techniques for utilizing fractals as therapeutic tools.

keywords          : "fractals, dwell time, growth sequence, decay sequence"

@@

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = F,
                      message = F)

@@

library(janitor)
library("papaja")
r_refs("r-references.bib")
# Sys.setenv(LANG = "en")

library(here)
library(rio)
library(knitr)
library(tidyverse)

theme_set(theme_minimal())
```

# Introduction

  Fractal is a term originally derived to describe the “roughness,” [@mandelbrot1989fractal] of objects, surfaces, or textures. Fractal properties are pervasive throughout nature [@losa2009fractal] and can be identified in the texture of clouds and mountains, in the branching pattern of trees and lightning, in the meandering shape of rivers and blood vessels. These fractal properties are identified by their self-similarity and scale invariance. Self-similarity implies that a part of the structure maintains the same structure or pattern as the whole. For example, if one were to zoom in on a small stretch of coastline, the jagged edges would look similar to the jaggedness of the coastline from a broader view. Likewise, a small section of a fern is nearly identical to the branching shape of the entire fern.

```{r, fig.show = "hold", out.width = "30%",fig.align = 'center'}
include_graphics(here("paper","paper_images","coast.jpg"))
include_graphics(here("paper","paper_images","fractalfern.png"))
include_graphics(here("paper","paper_images","spooky.png"))
```

  Scale-invariance is a related mathematical property which can be applied to objects that do not change across different scales; there is a similar patterm at all levels of magnification. Expressed more intuitively, scale invariance implies that it does not matter at which scale one views and object or observes a process, because the object or process contains a commonality at all scales.

@@

  In addition to inventing the term fractal, Mandelbrot developed a mathematics and geometry of fractals. He devised a way to model and generate complex systems with a simple mathematical rule that could be run repeatedly or iterated on itself. These iterative processes essentially use a rule to take an initial condition or state, change or modulate it, then put the resulting updated state back into the rule to generate another new state. This can be repeated infinitely and often results in high complex objects or systems. Fractal mathematics have continued to be developed and utilized in mathematics, physics, biology, chemistry, and engineering.A recent application of fractal mathematics is utilized in computer generated images and models. By starting with a simple algorithm, a mathematical equation or rule, graphic artists and programmers can create highly complex models of natural systems.

@@

  Two premiere examples of these mathematical or statistical fractals can be seen in the Koch Curve and the Cantor Set. In each case an image is created by starting with a single line segment and a rule for changing that line segment. The cantor set can be visualized by displaying each cycle of the process of iteration together. First a single line segment is displayed. The Cantor Set rule divides the segment in three equal sections and removes the middle section, leaving two line segments each one third of the original length and separated by a space equal to one third of the original line segment. The two separated line segments of the 2nd order iteration can be displayed beneath the original first order line segment. The rule, dividing and removing a portion of the line segment is applied to the two new line segments. Each of these line segments is divided into thirds and the middle third is removed. This results in four new smaller line segments spaced in a proportion equal to the previous two line segments. These four line segments are again displayed under the two line segments. This process can be repeated infinitely many times, dividing and removing segments at each iteration. When displayed together, we are able to visualize the repeating pattern which relates to the simple unchanging rule. This is an intuitive way to represent the properties of self-similarity, scale invariance, and iterative growth rules of fractals. A similar example is the Koch Curve. Again we start with a single line segment, but define a new rule to change that line segment through each iteration. Like the Cantor set, the line segment is divided into thirds, but instead of just removing the middle third which results in two new line segments, we bridge the middle third with two line segments that compose two sides of an isosceles triangle which spans the removed middle third of the original line. This new segmented line of the 2nd order iteration appears as a line with a triangular bump or protrusion in the middle. The entire segment is composed of four connected line segments. Again, by using the same algorithmic rule, divide a segment into thirds and span the removed middle third with an isosceles angle, and is applied to all four of the line segments created by the first iterative change. Now each of these four line segments will include an angular bump or peak that spans their middle third.These results in a set of 16 angular connected line segments. Just as we did with the cantor set, and can do with all fractal growth iterations, we can continue applying the rule to the updated condition of the object. This time, unlike the cantor set which displays all the iterations together, a single, new object is produced at each iteration. We can visualize the growth process by displaying a few of these objects together.

```{r, fig.show = "hold", out.width = "50%",fig.align = 'center'}
include_graphics(here("paper","paper_images","bars.png"))
include_graphics(here("paper","paper_images","star.jpg"))
```

  There are numerous examples of these types of geometric transformations which result in beautiful and striking constructions. Some well known fractal geometries are known as, The Dragon Curve, Sierpinski Triangle, and Hilbert Curve. These examples show the diversity of implementations of fractal properties in geometry. The variety of initial conditions and algorithmic rules result in an astounding array of complex objects which have real world applications and instantiations.

```{r, fig.show = "hold", out.width = "30%",fig.align = 'center'}
include_graphics(here("paper","paper_images","squigly.jpg"))
include_graphics(here("paper","paper_images","triangle.jpg"))
include_graphics(here("paper","paper_images","maze.jpg"))
```

@@

```{r, out.height = "30%",fig.align = 'center'}
include_graphics(here("paper","paper_images","morefractals.png"))
include_graphics(here("paper","paper_images","square.png"))
```

@@

  To date there have been numerous research studies which explore the aesthetic preferences for fractals of different dimensions. Studies have included methods which present participants with a two alternative forced choice task of two fractals of differing dimension. Participants select the fractal image they prefer aesthetically and after viewing many pairs with differing dimensions, statistics are used to calculate the most preferred dimensionality range across participants. The overarching results of previous research have shown that humans, both adults and children prefer fractals within a range of dimensions that coincide with the fractal dimension that is found in natural objects and scenes, usually between 1.3 and 1.5.Remembering that fractal dimension is an analog for complexity, we can conclude that most humans prefer a moderate level of complexity compared to low or high complexity near the extremes. A secondary analysis of fractal fluency has shown that humans often also prefer fractal images to non-fractal images and that we prefer fractals which more closely resemble natural images to fractals which are geometrically exact. [@spehar2003universal] Extensions on these research findings have led to the use of fractal images in therapeutic applications. Studies have shown some benefits of experiencing fractals, including physiological responses in EEG measurement of brain activity while viewing fractals [@hagerhall2015human], the reduction of stress using fractal art and architecture [@taylor2006reduction] and even in the experience of hearing fractal melodies [@beauvois2007quantifying]. There appears to be a clear link between, the fractal’s representation of complexity, the connection with natural objects and processes, and the human mind.

@@

```{r, fig.show = "hold", out.width = "30%",fig.align = 'center'}
include_graphics(here("paper","paper_images","blotches.png"))
include_graphics(here("paper","paper_images","moreblotches.png"))
include_graphics(here("paper","paper_images","deconstruction.png"))
```

@@

  Participants viewed self paced slideshows of fractal images in sequences of iterative growth and decay phases as well as a random ordering. The time spent viewing each image was measured as a Dwell Time, following the procedures of Baldwin et al. 2019. A set of images were generated for each of fourteen different fractal types iterated from 4 to 9 times. Each participant viewed a sequential slideshow and a random slideshow. Sequential slideshows presented each fractal type through phases of growth and decay from the least complex to most complex iterations and back down from most complex to least complex. Random slideshows presented the entire set of fractal images in random order. Six participants viewed Sequential Slideshow A, starting with Fractal Type 1 to Type 14, and Random Slideshow A starting with Fractal Image 1 to Image 174. The remaining 10 participants viewed Sequential Slideshow B, starting with Fractal Type 14 to Type 1, and Random Slideshow B starting with Fractal Image 174 to Image 1. This counterbalancing was included to address the findings of previous dwell time studies in which participants dwell times systematically decrease through the slideshow. In this way, the same fractal types were not always seen at the beginning or end of the slideshow. This balancing was then reversed in analysis so that each fractal type and the corresponding dwell times could be analyzed for the entire set of 16 participants together. Fractal images were generated using Wolfram Mathematica and slideshows were programmed using PsychoPy 3.

```{r, fig.show = "hold", out.height = "25%",fig.align = 'center'}
include_graphics(here("paper","paper_images","task_outline.jpg"))
include_graphics(here("paper","paper_images","fractal_direction.png"))
```

## Participants

The study recruited 16 healthy young adults ages 16-35 through the UO SONA recruitment system. All participants had normal or corrected to normal vision with no history of neurological damage.

## Data analysis

We used `r cite_r("r-references.bib")` for all our analyses.

@@

```{r import_data}
fractal_data <- clean_names(import(here("data", "frac_total.csv")))

raw_A <- clean_names(import(here("data", "frac_SA.csv")))

raw_B <- clean_names(import(here("data", "frac_SB.csv")))
```

```{r clean_data}
fractal_data <- fractal_data %>% 
  pivot_longer(cols = a:p,
               names_to = "participant",
               values_to = "dwell_time")

fractal_data <- fractal_data %>% 
  mutate(direction = factor(direction, labels = c("Decay", "Growth", "Random")),
         participant = factor(participant),
         disp_image = factor(disp_image))

fractal_data <- fractal_data %>% 
  mutate(fractal_type = case_when(
    grepl("Fern_A", disp_image, fixed = T) ~ "Fern_A",
    grepl("Fern_B", disp_image, fixed = T) ~ "Fern_B",
    grepl("Fractal_Curves_A", disp_image, fixed = T) ~ "Fractal_Curves_A",
    grepl("Fractal_Curves_B", disp_image, fixed = T) ~ "Fractal_Curves_B",
    grepl("Fractal_Curves_C", disp_image, fixed = T) ~ "Fractal_Curves_C",
    grepl("Fractal_Curves_D", disp_image, fixed = T) ~ "Fractal_Curves_D",
    grepl("Fractal_Curves_E", disp_image, fixed = T) ~ "Fractal_Curves_E",
    grepl("Hilbert", disp_image, fixed = T) ~ "Hilbert",
    grepl("Moore", disp_image, fixed = T) ~ "Moore",
    grepl("Pythagoras_A", disp_image, fixed = T) ~ "Pythagoras_A",
    grepl("Pythagoras_B", disp_image, fixed = T) ~ "Pythagoras_B",
    grepl("Sierpinski_A", disp_image, fixed = T) ~ "Sierpinski_A",
    grepl("Sierpinski_B", disp_image, fixed = T) ~ "Sierpinski_B",
    grepl("Tree_A", disp_image, fixed = T) ~ "Tree_A"
  ))

fractal_data <- fractal_data %>% 
  mutate(sequence_position = case_when(
    grepl("A0.png", disp_image, fixed = T) ~ 1,
    grepl("B0.png", disp_image, fixed = T) ~ 1,
    grepl("C0.png", disp_image, fixed = T) ~ 1,
    grepl("D0.png", disp_image, fixed = T) ~ 1,
    grepl("E0.png", disp_image, fixed = T) ~ 1,
    grepl("_0.png", disp_image, fixed = T) ~ 1,
    grepl("A100.png", disp_image, fixed = T) ~ 2,
    grepl("B100.png", disp_image, fixed = T) ~ 2,
    grepl("A1.png", disp_image, fixed = T) ~ 2,
    grepl("B1.png", disp_image, fixed = T) ~ 2,
    grepl("C1.png", disp_image, fixed = T) ~ 2,
    grepl("D1.png", disp_image, fixed = T) ~ 2,
    grepl("E1.png", disp_image, fixed = T) ~ 2,
    grepl("_1.png", disp_image, fixed = T) ~ 2,
    grepl("A200.png", disp_image, fixed = T) ~ 3,
    grepl("B200.png", disp_image, fixed = T) ~ 3,
    grepl("A2.png", disp_image, fixed = T) ~ 3,
    grepl("B2.png", disp_image, fixed = T) ~ 3,
    grepl("C2.png", disp_image, fixed = T) ~ 3,
    grepl("D2.png", disp_image, fixed = T) ~ 3,
    grepl("E2.png", disp_image, fixed = T) ~ 3,
    grepl("_2.png", disp_image, fixed = T) ~ 3,
    grepl("A400.png", disp_image, fixed = T) ~ 4,
    grepl("B400.png", disp_image, fixed = T) ~ 4,
    grepl("A3.png", disp_image, fixed = T) ~ 4,
    grepl("B3.png", disp_image, fixed = T) ~ 4,
    grepl("C3.png", disp_image, fixed = T) ~ 4,
    grepl("D3.png", disp_image, fixed = T) ~ 4,
    grepl("E3.png", disp_image, fixed = T) ~ 4,
    grepl("_3.png", disp_image, fixed = T) ~ 4,
    grepl("A800.png", disp_image, fixed = T) ~ 5,
    grepl("B800.png", disp_image, fixed = T) ~ 5,
    grepl("A4.png", disp_image, fixed = T) ~ 5,
    grepl("B4.png", disp_image, fixed = T) ~ 5,
    grepl("C4.png", disp_image, fixed = T) ~ 5,
    grepl("D4.png", disp_image, fixed = T) ~ 5,
    grepl("E4.png", disp_image, fixed = T) ~ 5,
    grepl("_4.png", disp_image, fixed = T) ~ 5,
    grepl("A1600.png", disp_image, fixed = T) ~ 6,
    grepl("B1600.png", disp_image, fixed = T) ~ 6,
    grepl("A5.png", disp_image, fixed = T) ~ 6,
    grepl("B5.png", disp_image, fixed = T) ~ 6,
    grepl("C5.png", disp_image, fixed = T) ~ 6,
    grepl("D5.png", disp_image, fixed = T) ~ 6,
    grepl("E5.png", disp_image, fixed = T) ~ 6,
    grepl("_5.png", disp_image, fixed = T) ~ 6,
    grepl("A3200.png", disp_image, fixed = T) ~ 7,
    grepl("B3200.png", disp_image, fixed = T) ~ 7,
    grepl("A6.png", disp_image, fixed = T) ~ 7,
    grepl("A6400.png", disp_image, fixed = T) ~ 8,
    grepl("B6400.png", disp_image, fixed = T) ~ 8,
    grepl("A7.png", disp_image, fixed = T) ~ 8,
    grepl("A12800.png", disp_image, fixed = T) ~ 9,
    grepl("B12800.png", disp_image, fixed = T) ~ 9
  ))

raw_A_long <- raw_A %>% 
 pivot_longer(cols = a:f, names_to = "participant", values_to = "dwell_time")

raw_B_long <- raw_B %>% 
 pivot_longer(cols = g:p, names_to = "participant", values_to = "dwell_time")
```

```{r exclude_outliers}
cutoff <- mean(fractal_data$dwell_time) + 3 * sd(fractal_data$dwell_time)

filtered_fractal_data <- fractal_data %>% 
  filter(dwell_time < cutoff)
```

```{r diff_dwell_time}
fractal_data_wide <- fractal_data %>% 
  pivot_wider(names_from = direction,
              values_from = dwell_time) %>% 
  select(-Random) %>% 
  mutate(Growth = unlist(Growth),
         Decay = unlist(Decay),
         diff_dwell_time = Growth - Decay)
```

```{r create_table}
table_byp <- fractal_data %>% 
  group_by(participant, direction) %>% 
  summarise(mean_dwell_time = round(mean(dwell_time), 3),
            sd_dwell_time = round(sd(dwell_time), 3))
```

```{r tbyp, results = "asis"}
apa_table(table_byp, col.names = c("Participant", "Sequence Direction", "Mean Dwell Time (s)", "SD Dwell Time (s)"), caption = "Descriptive statistics of dwell times by participant and direction", note = "Participant means and standard deviations for dwell times by sequence direction (growth, decay, or random). All units are in seconds (s).", placement = "htbp")
```

```{r plots}
new_data_for_now <- fractal_data %>% 
  group_by(participant, direction) %>% 
  summarise(mean_dwell_time = round(mean(dwell_time), 3),
            sd_dwell_time = round(sd(dwell_time), 3))

new_data_for_later <- fractal_data %>% 
  group_by(direction) %>% 
  summarise(mean_dwell_time = round(mean(dwell_time), 3),
            sd_dwell_time = round(sd(dwell_time), 3),
            mean_se_dwell_time = round(plotrix::std.error(dwell_time), 3))

ymin = new_data_for_later$mean_dwell_time - new_data_for_later$mean_se_dwell_time
ymax =new_data_for_later$mean_dwell_time + new_data_for_later$mean_se_dwell_time
```

  Descriptive statistics for each participant by direction of sequence are shown in Table\ \@ref(tab:tbyp).

```{r main, fig.cap = "Error bars represent standard errors."}
new_data_for_later %>% 
  ggplot(aes(x = direction, y = mean_dwell_time, fill = direction)) +
  geom_col() +
  labs(x = "Direction", y = "Dwell Time (s)", title = "Dwell Times by Condition") +
  theme(legend.position = "none") +
  geom_errorbar(aes(ymin = ymin, ymax = ymax, width = .5)) +
  scale_fill_viridis_d()
```

```{r main_model}
model1 <- lm(data = fractal_data, dwell_time ~ direction)

anova1 <- car::Anova(model1)

anova1a <- apa_print(anova1)
```

 A one-way ANOVA was conducted to examine the effect of fractal presentation directions on dwell times. There was a significant effect of direction on dwell time, `r anova1a$full_result$direction`. These results demonstrate that growth conditions showed significantly longer dwell times compared to the decay or random directional conditions. These results are depicted in Figure\ \@ref(fig:main).

```{r fractal_type}
seq_fractal_data <- fractal_data %>% 
  filter(direction != "Random") %>% 
  mutate(sequence_position = factor(sequence_position),
         fractal_type = factor(fractal_type))

new_data_forever <- seq_fractal_data %>% 
  group_by(sequence_position, fractal_type) %>% 
  summarise(mean_dwell_time = round(mean(dwell_time), 3),
            sd_dwell_time = round(sd(dwell_time), 3))

model2 <- lm(data = seq_fractal_data, dwell_time ~ sequence_position * fractal_type)

anova2 <- car::Anova(model2)

anova2a <- apa_print(anova2)

temp <- seq_fractal_data %>% 
  mutate(seq_pos2 = case_when(
    direction == "Decay" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 1 ~ 18,
    direction == "Decay" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 2 ~ 17,
    direction == "Decay" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 3 ~ 16,
    direction == "Decay" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 4 ~ 15,
    direction == "Decay" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 5 ~ 14,
    direction == "Decay" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 6 ~ 13,
    direction == "Decay" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 7 ~ 12,
    direction == "Decay" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 8 ~ 11,
    direction == "Decay" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 9 ~ 10,
    direction == "Growth" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 1 ~ 1,
    direction == "Growth" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 2 ~ 2,
    direction == "Growth" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 3 ~ 3,
    direction == "Growth" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 4 ~ 4,
    direction == "Growth" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 5 ~ 5,
    direction == "Growth" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 6 ~ 6,
    direction == "Growth" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 7 ~ 7,
    direction == "Growth" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 8 ~ 8,
    direction == "Growth" & (fractal_type == "Fern_A" | fractal_type == "Fern_B") & sequence_position == 9 ~ 9,
    direction == "Decay" & (fractal_type == "Fractal_Curves_A" | fractal_type == "Fractal_Curves_B" | fractal_type == "Fractal_Curves_C" | fractal_type == "Fractal_Curves_D" | fractal_type == "Fractal_Curves_E" | fractal_type == "Hilbert" | fractal_type == "Moore" | fractal_type == "Sierpinski_B") & sequence_position == 1 ~ 12,
    direction == "Decay" & (fractal_type == "Fractal_Curves_A" | fractal_type == "Fractal_Curves_B" | fractal_type == "Fractal_Curves_C" | fractal_type == "Fractal_Curves_D" | fractal_type == "Fractal_Curves_E" | fractal_type == "Hilbert" | fractal_type == "Moore" | fractal_type == "Sierpinski_B") & sequence_position == 2 ~ 11,
    direction == "Decay" & (fractal_type == "Fractal_Curves_A" | fractal_type == "Fractal_Curves_B" | fractal_type == "Fractal_Curves_C" | fractal_type == "Fractal_Curves_D" | fractal_type == "Fractal_Curves_E" | fractal_type == "Hilbert" | fractal_type == "Moore" | fractal_type == "Sierpinski_B") & sequence_position == 3 ~ 10,
    direction == "Decay" & (fractal_type == "Fractal_Curves_A" | fractal_type == "Fractal_Curves_B" | fractal_type == "Fractal_Curves_C" | fractal_type == "Fractal_Curves_D" | fractal_type == "Fractal_Curves_E" | fractal_type == "Hilbert" | fractal_type == "Moore" | fractal_type == "Sierpinski_B") & sequence_position == 4 ~ 9,
    direction == "Decay" & (fractal_type == "Fractal_Curves_A" | fractal_type == "Fractal_Curves_B" | fractal_type == "Fractal_Curves_C" | fractal_type == "Fractal_Curves_D" | fractal_type == "Fractal_Curves_E" | fractal_type == "Hilbert" | fractal_type == "Moore" | fractal_type == "Sierpinski_B") & sequence_position == 5 ~ 8,
    direction == "Decay" & (fractal_type == "Fractal_Curves_A" | fractal_type == "Fractal_Curves_B" | fractal_type == "Fractal_Curves_C" | fractal_type == "Fractal_Curves_D" | fractal_type == "Fractal_Curves_E" | fractal_type == "Hilbert" | fractal_type == "Moore" | fractal_type == "Sierpinski_B") & sequence_position == 6 ~ 7,
    direction == "Growth" & (fractal_type == "Fractal_Curves_A" | fractal_type == "Fractal_Curves_B" | fractal_type == "Fractal_Curves_C" | fractal_type == "Fractal_Curves_D" | fractal_type == "Fractal_Curves_E" | fractal_type == "Hilbert" | fractal_type == "Moore" | fractal_type == "Sierpinski_B") & sequence_position == 1 ~ 1,
    direction == "Growth" & (fractal_type == "Fractal_Curves_A" | fractal_type == "Fractal_Curves_B" | fractal_type == "Fractal_Curves_C" | fractal_type == "Fractal_Curves_D" | fractal_type == "Fractal_Curves_E" | fractal_type == "Hilbert" | fractal_type == "Moore" | fractal_type == "Sierpinski_B") & sequence_position == 2 ~ 2,
    direction == "Growth" & (fractal_type == "Fractal_Curves_A" | fractal_type == "Fractal_Curves_B" | fractal_type == "Fractal_Curves_C" | fractal_type == "Fractal_Curves_D" | fractal_type == "Fractal_Curves_E" | fractal_type == "Hilbert" | fractal_type == "Moore" | fractal_type == "Sierpinski_B") & sequence_position == 3 ~ 3,
